The Small World Network Effect in Software Project Teams \author{Kevin Peterson\\ \small \texttt{pete1968@umn.edu} }

### 2013-11-26 paper/interview1.tex
Updated code to display true Q value.

\subsection*{Q1: How do development teams collaborate and share ideas?}

\subsection*{Q2: Do teams generally remain intact over multiple projects, or do members often shuffle and re-combine?}

\subsection*{Q3: When assembling a team, which is preferred -- members familiar with each other, or members familiar with the task or project.}

\subsection*{Q4: What are some characteristics of a high performing team?}

\subsection*{Q5: In your experience, is it better to (A) keep an innovative team intact, (B) periodically bring in new members or re-assign existing members, or (C) split up and disperse the team to as many other projects as possible?}

>>>>

\subsection*{Q1: How do development teams collaborate and share ideas?}
Co-location is important for teams, as it really fosters communication and collaboration. Being able to quickly talk with someone and resolve a problem is very important. In some cases, having the team in the same room is beneficial, especially if the project is going badly. That way, ideas can get discussed and problems resolved very quickly. Regular meetings are also important with large teams. With these meetings, it is important to have the correct tools to facilitate communication. These meetings help projects have visibility into each other. Management support is essential to this communication process. Meetings between teams can consist of all members of the teams if the teams are small enough, but most of the time should just be a representative.

\subsection*{Q2: Do teams generally remain intact over multiple projects, or do members often shuffle and re-combine?}
I have had both. It really depends on the projects and the situation at the time. It can be nice to have new people brought in to make things different, and to bring some new thoughts and ideas to the team. Even one person can change the dynamics of a team, either positively or negatively. As a project manager, I have to be watchful of this, as negativity can spread quickly. I have also noticed that as people move around to different projects, they are generally brought up to the level of the team. For example, if somebody joins a high performing team, they tend to elevate their skill to match that of the group, which is a good thing.

\subsection*{Q3: When assembling a team, which is preferred -- members familiar with each other, or members familiar with the task or project.}
Familiarity is good, but you have to account for burnout if people work on the same project for too long. This really depends on project dynamics, special skills needed, etc. It is important to note, however, that team members can be familiar with each other but not get along well, or not work well with each other. I have had projects where team members were very familiar with each other, but did not function well as a team. Clashing personalities and other team problems can cause a toxic environment. Most importantly, whether the team is familiar with each other or not, is that they can share knowledge and skills.

\subsection*{Q4: What are some characteristics of a high performing team?}
Most teams I notice go through the Forming, Storming, Norming and Performing stages. Good teams usually have a strong leader, a good vision, and good management support. Of course, they also possess the necessary skills for the project (or, if not, are willing to learn). Good teams have respect for each other, their leadership, and their management. Professionalism is also an indicator of a high performing team. Teams that truly care about their job, the team, and the project are generally more successful. Teams are generally more successful if the individuals themselves are skilled and talented. Being able to react to change is important, and good teams generally have good change management procedures.

\subsection*{Q5: In your experience, is it better to (A) keep an innovative team intact, (B) periodically bring in new members or re-assign existing members, or (C) split up and disperse the team to as many other projects as possible?}
In general, I like to keep good teams intact. It is, however, nice to bring in or add a new member to a team. One strategy I like to employ is to bring in a new member, train them and let them learn from the high performing team, and move them out of the project and into a new one. This keeps the core team intact, which is important. Priorities can greatly influence this, and sometimes personnel moves are driven from higher management priorities. If experts are removed from a high performing team, the team may suffer. This is, however, a risk that knowledge becomes concentrated. High priority/high visibility teams have a tendency to get the highest performing individuals, which can cause other projects to suffer.

### 2013-11-26 paper/main.tex
Added start of interview.
Updated code to display true Q value.

\section{Introduction}
A social network is a graph people and the interactions between them. The dynamics of social networks have proved to be an important research topic for many different areas of study, from academic publication\cite{barabasi2002evolution}, to the success of broadway musicals\cite{uzzi2005collaboration}. In general, a social network is important to understanding how ideas and influence are spread\cite{kempe2003maximizing}, from which we can investigate the best conditions for optimal group perfomance within these networks.

A `small world' network is a social network characterized by high clustering of graph nodes paired with a short average path between nodes \cite{watts1998collective}. `Clustering' in this context means how closely nodes in the graph are related to each other, or "the friend of my friend is also my friend." Milgram, in is seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}. From these observations, hypotheses can be made as to the efficiency of how knowledge is transfered in these groups\cite{latora2001efficient}. In this work, we investigate the `small world' phenomenon and how it relates to Software Project Teams.

@@

\subsection{Collection and Storage}
Data was collected using the freely available FLOSSmole\cite{floss2006} data sets of free and open source projects. The data collection used was collected from the popular open source project site FreeCode\footnote{http://freecode.com/}, and contains data up to September, 2013. Results were analyzed using the Python graph processing packaged NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the projects and the edges representing one or more shared developer between projects. Various properties of the graph are analyzied, as indicated in the sections below.

\subsubsection{Graph Structure}
We first consider the FreeCode FLOSSmole\cite{floss2006} data sets to be one large graph. From this large disconnected graph, each connected subgraph is analyzed. Metrics are computed for each of these subgraphs, and a total subgraph performance is computed by averaging the performance value of each graph in the subgraph. We measure a subgraph's perfomance by the FreeCode \textit{Popularity Score} metric, which is calculated the following formula\footnote{http://help.freecode.com/kb/statistics/how-do-you-measure-a-projects-popularity}:

@@

A bipartite graph is a specialized type of graph containing two disjoint sets of verticies. This can be denoted by ${G=(P,C,E)}$, where ${P \cap C = \emptyset}$. For our purposes, let set $P$ be the set of all projects, and set $C$ be the set of all contributors. Each project node, therefore, is only connected to contributor nodes, and vice versa, as show in figure \ref{fig:example_bipartite_graph}. This bipartite representation is typical of what is found in social and collaboration networks\cite{ramasco2004self}.

A bipartite graph projection. in necessary to further analyze the graph. A bipartite projection involves taking the disjoint node sets $P$ and $C$, and representing the graph as relationships between only one of those sets of nodes. A projection onto the project nodes, or $P$, of figure \ref{fig:example_bipartite_graph} is show in figure \ref{fig:example_bipartite_projection_graph}. Here, projects are connected directly, with edges occurring if one or more developer is shared between the projects. It is important to note that the projection could have been done in terms of the contributors, not the projects. This would have led to a graph with nodes of contributors $C$, each being linked by sharing a common project. This approach is another common way of representing these bipartite graphs\cite{newman2001scientific}, and illustrates the fact that a bipartite projections are lossy compared to the data reprented in the original graph\cite{zhou2007bipartite}.\\

@@

\noindent\\\textit{Clustering Coefficient - Transitivity ($C^\Delta$)}\\
We calculate the \textit{Clustering Coefficient} (or $C^\Delta$) as a measure of the proportion of closed triangles in a graph\cite{newman2003structure}:
\[C^\Delta = 3 \times \frac{\text{number of triangles}}
                    {\text{number of connected triples}}\]

\noindent\\\textit{Clustering Coefficient - Average ($\overline{C^\Delta}$)}\\
Another approach to clustering is to take the average of each node's local clustering coefficient\cite{watts1998collective}.

@@

\noindent\\\textit{Average Path Length ($L$)}\\
Also, calculating the \textit{Path Length} ($L$) allows us to determine how far removed developers in the graph are from one another. The path length is calculated as the average distance between any two node pairs in the graph:

@@

\noindent\\\textit{Small World Quotient ($Q$)}\\
The \textit{Small World Quotient} ($Q$) is defined as the Clustering Coefficient ($C^\Delta$) divided by the  Average Path Length ($L$)\cite{watts1999small,watts1998collective}.

@@

@@

\noindent\\\textit{Small-World-ness ($S^\Delta$)}\\
The definition of a \textit{small world} graph can be formalized by the following equations\cite{humphries2008network}.

@@

\noindent\\\textit{Successful vs Unsuccessful}\\
To further explore the data, a comparion was made between \textit{successful} and \textit{unsuccessful} projects was made. We define \textit{successful} as any project with {$PopularityScore > 100$}, and \textit{unsuccessful} as any project with {$PopularityScore < 50$}. Given this view of the data, we can then make assertions about how we expect these types of projects to behave, given the stated hypothesis. More specifically, we can examine not only if the u-shaped curve holds for \textit{popular} projects, but if the inverse holds for \textit{unsuccessful} ones.

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.5\textwidth]{images/freecode-smallworld-histo.png}
\end{center}
\caption{$S^\Delta$ Distribution}
\label{fig:sw_distribution}
\end{figure}

@@

>>>>

\section{Introduction}
A social network is a graph of people and the connections between them. The dynamics of social networks have proved to be an important research topic for many different areas of study, from academic publication\cite{barabasi2002evolution}, to the success of broadway musicals\cite{uzzi2005collaboration}. In general, a social network is important to understanding how ideas and influence are spread\cite{kempe2003maximizing}, from which we can investigate the best conditions for optimal group perfomance within these networks.

A `small world' network is a social network characterized by high clustering of graph nodes paired with a short average path between nodes\cite{watts1998collective}. `Clustering' in this context means how closely nodes in the graph are related to each other, or "the friend of my friend is also my friend." Milgram, in is seminal study, found that any two people are linked through a chain of friends and acquaintances on average of 6 people long\cite{milgram1967small}, or in other words, the average `path length' between any two people.

Uzzi \cite{uzzi2005collaboration} expanded on this by exploring how these small world networks correlate to collaboration and creativity. In his study, the network of artists involved in the making of Broadway musicals from 1945 to 1989 were studied. Performance of the musicals were then measured based on financial success and critical acclaim, and compared that quantitative analysis to the clustering and path length of the graph.

We follow Uzzi's work, and expand it to software teams. We hypothesize that the qualities of a small world graph that foster collaboration, creativity, and the efficiency of how knowledge is transfered\cite{latora2001efficient} will also apply to software project teams. In this work, we investigate the `small world' phenomenon quantitatively by exploring open source project data, and qualitatively by a series of interviews of subject matter experts, and look to compare Uzzi's findings to ours.

@@

\subsection{Collection and Storage}
Data was collected using the freely available FLOSSmole\cite{floss2006} data sets of free and open source projects. The data collection used was collected from the popular open source project site FreeCode,\footnote{http://freecode.com/} and contains data up to September, 2013. Results were analyzed using the Python graph processing packaged NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

\subsection{Analysis}
The collected data was organized into a graph structure, with the nodes representing the projects and the edges representing one or more shared developer between projects. Various properties of the graph are analyzed, followed by a qualitative section to help link the quantitative analysis to subject matter expert experiences.

\subsubsection{Graph Structure}
We first consider the FreeCode FLOSSmole\cite{floss2006} data sets to be one large graph. From this large disconnected graph, each connected subgraph is analyzed. Metrics are computed for each of these subgraphs, and a total subgraph performance value is computed by averaging the performance value of each graph in the subgraph. We measure a subgraph's performance by the FreeCode \textit{Popularity Score} metric, which is calculated the following formula\footnote{http://help.freecode.com/kb/statistics/how-do-you-measure-a-projects-popularity}:

@@

The FreeCode data can be represented as a bipartite graph. A bipartite graph is a specialized type of graph containing two disjoint sets of verticies. This can be denoted by ${G=(P,C,E)}$, where ${P \cap C = \emptyset}$. For our purposes, let set $P$ be the set of all projects, and set $C$ be the set of all contributors. Each project node, therefore, is only connected to contributor nodes, and vice versa, as show in figure \ref{fig:example_bipartite_graph}. This bipartite representation is typical of what is found in social and collaboration networks\cite{ramasco2004self}.

A bipartite graph projection is necessary to further analyze the graph. A bipartite projection involves taking the disjoint node sets $P$ and $C$, and representing the graph as relationships between only one of those sets of nodes. A projection onto the project nodes, or $P$, of figure \ref{fig:example_bipartite_graph} is show in figure \ref{fig:example_bipartite_projection_graph}. Here, projects are connected directly, with edges occurring if one or more developer is shared between the projects. It is important to note that the projection could have been done in terms of the contributors, not the projects. This would have led to a graph with nodes of contributors $C$, each being linked by sharing a common project. This approach is another common way of representing these bipartite graphs\cite{newman2001scientific}, and illustrates the fact that a bipartite projections are lossy compared to the data reprented in the original graph\cite{zhou2007bipartite}.\\

@@

\noindent\\\textit{\textbf{Clustering Coefficient - Transitivity ($C^\Delta$)}}\\
We calculate the \textit{Clustering Coefficient} (or $C^\Delta$) as a measure of the proportion of closed triangles in a graph\cite{newman2003structure}:
\[C^\Delta = 3 \times \frac{\text{number of triangles}}
                    {\text{number of connected triples}}\]

This metric shows us how closely related, or ``clustered,'' that the graph is. It is also the probability that two nodes will be connected if they share a common neighbor\cite{newman2003properties}. In our context, if project $P_1$ and $P_2$ share a developer with $P_3$, this metric represents the probability that $P_1$ and $P_2$ will share a developer.

\noindent\\\textit{\textbf{Clustering Coefficient - Average ($\overline{C^\Delta}$)}}\\
Another approach to clustering is to take the average of each node's local clustering coefficient\cite{watts1998collective}.

@@

\noindent\\\textit{\textbf{Average Path Length ($L$)}}\\
Also, calculating the \textit{Path Length} ($L$) allows us to determine how far removed developers in the graph are from one another. The path length is calculated as the average distance between any two node pairs in the graph:

@@

\noindent\\\textit{\textbf{Small World Quotient ($Q$)}}\\
The \textit{Small World Quotient} ($Q$) is defined as the Clustering Coefficient ($C^\Delta$) divided by the  Average Path Length ($L$)\cite{watts1999small,watts1998collective}.

@@

Calculating $Q$ allows us to

@@

\noindent\\\textit{\textbf{Small-World-ness ($S^\Delta$)}}\\
The definition of a \textit{small world} graph can be formalized by the following equations\cite{humphries2008network}.

@@

\noindent\\\textit{\textbf{Successful vs Unsuccessful}}\\
To further explore the data, a comparion was made between \textit{successful} and \textit{unsuccessful} projects was made. We define \textit{successful} as any project with {$PopularityScore > 100$}, and \textit{unsuccessful} as any project with {$PopularityScore < 50$}. Given this view of the data, we can then make assertions about how we expect these types of projects to behave, given the stated hypothesis. More specifically, we can examine not only if the u-shaped curve holds for \textit{popular} projects, but if the inverse holds for \textit{unsuccessful} ones.

@@

@@

\input{interview1}
\input{interview2}
